A Bidirected String Graph Model for Genome Assembly

### 2012-12-11 presentation/paper.tex
paper: Implementation, more citations

\documentclass[10pt]{article}
\usepackage[left=3.3cm,right=3.3cm,bottom=3.0cm,top=3.0cm]{geometry}
\usepackage{float}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{comment}
\usepackage{multicol}
%\usepackage{amsthm}

@@

Sequencing an organism's genome is an important step to further genetic research
on that organism.  However, currently available sequencing technology can only
sequence short fragments of DNA.  As a result, the whole-genome shotgun
sequencing approach to genome sequencing relies on breaking up a sample of
genomic DNA into fragments, sequencing those fragments, then using a program
called a {\em genome assembler} to put together these short sequences to
reconstruct the original, much longer sequence.  This paper provides an
introduction and formalization of the genome assembly problem, then introduces a
model and algorithm for genome assembly that is based on a special type of graph
called a \KeyTerm{bidirected string graph}.  \end{abstract}

@@

The running time of the seed-and-extend algorithm, as described above, is
typically much faster than the naive algorithm, although it is highly dependent
on the data.  For a string of the genome of length $G$ and with a seed length
$k$, there will be at most $G-k+1$ canonical seeds.\footnote{The canonical form
of a DNA sequence and its reverse-complement is the lexicographically lesser of
the two.  Here, we refer canonical form of each seed, because we wish to compare
reads that share a seed in either a forward or reverse-complement manner.}  For
each of these seeds, the reads containing it will be examined exhaustively in a
pairwise manner, which takes quadratic time with respect to the number of reads
sharing that seed.  Since the more densely reads are sampled from the genome,
the more reads will contain each seed,  the running time of the seed-and-extend
algorithm rises quadratically with regards to read density.  In addition, the
number of read pairs the algorithm considers for a specific seed also rises
quadratically with the number of times that seed is repeated in the genome due
to genomic repeats.  Overall, the seed-and-extend algorithm is not ideal and is
still a bottleneck of the entire assembly algorithm, despite the fact that it is
much faster than the naive algorithm (except in degenerate cases, such as a
``genome'' that is entirely a single repeated letter).

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=0.6\textwidth]{E_coli.pdf}
		\caption{Transitively-reduced, collapsed bidirected string graph for
		1000 bp perfect reads uniformly sampled every 100 bp from random strands
		of the non-circular version of the 4,639,675 bp genome of {\it E. coli}
		(a common bacterium) using a minimum overlap length of 250 bp.  This
		graph was produced by the code that has been discussed, but the actual
		drawing of the picture was done by {\it GraphViz}.  The 46,387 reads
		that were provided as input to the assembler were generated by a Perl
		script.  The bottleneck of the algorithm up to this point is
		\ProgramName{compute-overlaps}, which took about 10 minutes and used a
		maximum of about 1 GiB of memory to find all 336,227 overlaps from these
		reads, seeding on 128-mers.  The running time and memory usage of
		\ProgramName{compute-overlaps} would need to improved further to handle
		larger genomes.  (Also it still does not handle inexact overlaps, which
		are not considered in this paper.)}
		\label{fig:E_coli}
	\end{center}
\end{figure}

@@

\section{Producing the final assembly}
\subsection{Overview}

@@

@@

\begin{itemize}
\item Finish the network flow and generalized Eulerian path parts of the algorithm
so that the implementation is ``complete'' and can produce a ``final'' assembly
\item Improve performance of \ProgramName{compute-overlaps}
\item Allow for sequencing errors by finding inexact overlaps and modifying other
stages of the algorithm, such as transitive reduction, to allow for the
possibility that overlaps are not exact.
\item Compensate for sequencing errors by attempting to correct them, either as
part of \ProgramName{compute-overlaps} or as a new stage of the algorithm.
\item Consider more recent publications that explore variations on this algorithm,
such as parellizing the algorithm on a cluster \cite{X}, or using the
Burrows-Wheeler transform to directly and efficiently construct the bidirected
string graph directly from the reads\cite{X}.
\item Use mate-pair information to help resolve repeats and connect assembled
sequences even when they are separated by unknown sequence
\end{itemize}

\section*{Acknowledgments}

Acknowledgements

>>>>

\documentclass[12pt]{article}
\usepackage[left=3.3cm,right=3.3cm,bottom=3.0cm,top=3.0cm]{geometry}
\usepackage{float}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{comment}
\usepackage{mathptmx}
%\usepackage{amsthm}

@@

Sequencing an organism's genome is an important step to further genetic research
on that organism.  However, currently available sequencing technologies can only
sequence short fragments of DNA.  As a result, the whole-genome shotgun
sequencing approach to genome sequencing relies on breaking up a sample of
genomic DNA into fragments, sequencing those fragments, then using a program
called a {\em genome assembler} to put together these short sequences to
reconstruct the original, much longer sequence.  This paper provides an
introduction and formalization of the genome assembly problem, then introduces a
model and algorithm for genome assembly that is based on a special type of graph
called a \KeyTerm{bidirected string graph}.  \end{abstract}

@@

The running time of this seed-and-extend algorithm is
typically much faster than the naive algorithm, although it is highly dependent
on the data.  The number of distinct seeds is at worst linear with respect to
the genome length.
%of a DNA sequence and its reverse-complement is the lexicographically lesser of
%the two.  Here, we refer canonical form of each seed, because we wish to
%compare reads that share a seed in either a forward or reverse-complement
%manner.}
For each of these distinct seeds, the reads containing it are examined pairwise,
which takes quadratic time with respect to the number of reads sharing the
seed.  Since more reads will contain each seed when the reads are sampled more
densely from the genome, the running time of the seed-and-extend algorithm rises
quadratically with respect to read density.  In addition, the number of read
pairs the algorithm considers for a specific seed also rises quadratically with
the number of times that seed is repeated in the genome due to genomic repeats.
Overall, the seed-and-extend algorithm is not ideal and is still a bottleneck of
the entire assembly algorithm, despite the fact that it is much faster than the
naive algorithm (except in degenerate cases, such as a ``genome'' that is
entirely a single repeated letter).

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=0.6\textwidth]{E_coli.pdf}
		\caption{Transitively-reduced, collapsed bidirected string graph for
		1000 bp perfect reads uniformly sampled every 100 bp from random strands
		of the non-circular version of the 4,639,675 bp genome of {\it E. coli}
		(a common bacterium) using a minimum overlap length of 250 bp.  This
		graph was produced by the code that has been discussed, but the actual
		drawing of the picture was done by {\it
		GraphViz}\cite{GraphVizPublication}.  The 46,387 reads
		that were provided as input to the assembler were generated by a Perl
		script.  The bottleneck of the algorithm up to this point is
		\ProgramName{compute-overlaps}, which took about 10 minutes and used a
		maximum of about 1 GiB of memory to find all 336,227 overlaps from these
		reads, seeding on 128-mers.  The running time and memory usage of
		\ProgramName{compute-overlaps} would need to improved further to handle
		larger genomes.  (Also it still does not handle inexact overlaps, which
		are not considered in this paper.)}
		\label{fig:E_coli}
	\end{center}
\end{figure}

@@

\section{Producing the final assembly}

\label{sec:final}

@@

\section{Implementation}

The implementation of the fragment string graph algorithm that was done for this
project is a set of C++ programs that iteratively transform the data on-disk
into the final assembly.  Each program takes as input one or more files as well
as zero or more other command-line arguments and produces as output one or more
files.  Each file contains data such as a list of reads, a list of overlaps, or
a graph.  Internally to the assembler pipeline, data is stored in binary form.
\Figure{fig:impl_graph} shows how the data flows through the main stages of the
assembler.

The input to the assembler is one or more files in plain-text form that each
contain a list of reads.  The final output of the assembler will eventually be a
file in plain-text format that gives the list of assembled contigs.  Currently,
the final stages of the algorithm (\Section{sec:final}, except for the
calculation of the A-statistics) have not been implemented as of the end of the
semester of Fall 2012.

\begin{figure}
	\begin{center}
		\includegraphics[width=0.7\textwidth]{programs.pdf}
	\end{center}
	\caption{The flow of data through the assembly pipeline.  Each node is a
	program, each arrow represents the output of a program being used as input
	in another program.  Yellow programs have not yet been implemented as of
	December 10, 2012.}
	\label{fig:impl_graph}
\end{figure}

@@

\begin{itemize}
\item Finish the network flow and generalized Eulerian path parts of the
algorithm so that the implementation is ``complete'' and can produce a ``final''
assembly.
\item Improve the performance of \ProgramName{compute-overlaps}.
\item Allow for sequencing errors by finding inexact overlaps and modifying other
stages of the algorithm, such as transitive reduction, to allow for the
possibility that overlaps are inexact.
\item Compensate for sequencing errors by attempting to correct them, either as
part of \ProgramName{compute-overlaps} or as a new stage of the algorithm.
\item Consider more recent publications that explore variations on this
algorithm, such as parallelizing the algorithm \cite{Jackson2008}, or using the
Burrows-Wheeler transform to efficiently construct the transitively-reduced
bidirected string graph directly from the reads\cite{Simpson2010}.
\item Use mate-pair information to help resolve repeats and connect assembled
sequences even when they are separated by unknown sequence, possibly using ideas
from \cite{Medvedev2009}.
\end{itemize}
